Site-specific substitution model inference with iterative tree likelihood maximization.

### 2018-12-28 notes/derivation.tex
updating notes

Using the product from of the likelihood, we can now write down compact expressions for the derivative of the likelihood and the with respect to the substitution model.
\begin{equation}
	\frac{\partial \log \LH}{\partial \mu^{b}} = \sum_{k\in T} \frac{t_k}{\mu^{b}}\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{equation}
Analogously, we find for the derivaties wrt to $\eqp^{k,b}_{i}$
\begin{equation}
\frac{\partial \log \LH}{\partial \eqp^{b}_i}= \sum_{k\in T} \frac{t_k}{\eqp^{b}_i}\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_i}{\eqp^b_i}
\end{equation}
where $\sigma^{0,b}_i$ corresponds to state of the root sequence at position $b$.
The corresponding derivative for the transition matrix $W_{rs}$ is
\begin{equation}
	\frac{\partial \log \LH}{\partial W_{ij}}= \sum_{k\in T} \frac{t_k}{W_{ij}} \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{equation}
The at an extremum, all these derivatives have to vanish.
In addition we need to impose the normalization constraints which is done in the form of Lagrange multipliers such that the final conditions are
\begin{eqnarray}
	0 &=& \sum_{k\in T} t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \nonumber \\
    0 &=& 	\sum_{k\in T} t_k\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}} + \sigma^{0,b}_i + \lambda^{b}\eqp^b_i + \kappa \sum_j W_{ij}\eqp_i^{b}\eqp_j^{b} \nonumber\\
	0 &=& \sum_{k\in T} t_k \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \kappa W_{ij}\sum_{b=1}^L \eqp^b_i\eqp^b_j
\end{eqnarray}
Summing the second and third equation over $i$ and $i,j$, respectively, and substituting the first equation, we find $\kappa=0$ and $\lambda^a=1$.

To arrive at an iterative scheme to solve these equations, it is helpful to split the individuals terms into \emph{diagonal} contributions (the case in which the sequence does not change along a branch) and \emph{off-diagonal} contributions (substitutions happened on that branch).
We will first consider the limit of short branches were we can approximate $e^{\mat{Q}^at}=\mat{I} + \mat{Q}^at$ and evaluate to leading order
\begin{equation}
\begin{split}
t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \approx &	t_k \frac{-\sum_{i} \mu^b\sum_h p_h^b W_{hi} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i, j\neq i} \mu^b\sigma^{k,b}_i p_i^bW_{ij} \rho^{k,b}_j}{\sum_{i} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i,j\neq i} \mu^b \sigma^{k,b}_i \eqp_i^b W_{ij}t_k \rho^{k,b}_j}\\
\end{split}
\end{equation}
In the limit of short branches (or if ancestral states are known), $\sigma^{k,b}_i$ and $\rho^{k,b}_i$ are either $0$ or $1$ such that only one of the two terms in the numerator and denominator contribute.
In case $\sigma^{k,b}_i=\rho^{k,b}_i$ (no substitution on that branch), the denominator is 1 and the numerator $- \mu^b \sum_{i,h}p_h^b W_{hi}t_k$.
If a substitution happened, numerator and denominator cancel with result 1.
After summing over all branches, the condition for the mutation rate simplifies to
\begin{equation}
	\mu^b \sum_i \eqp_i^b W_{ij}T_j^b = \sum_{i\neq j} n_{ij}
\end{equation}
where $T_j^b$ is the sum of all $t_k$ were the sequence is in state $j$.

>>>>

Using the product from of the likelihood, we can now write down compact expressions for the derivative of the likelihood and the with respect to the substitution model.
\begin{equation}
	\frac{\partial \log \LH}{\partial \mu^{b}} = \sum_{k\in T} \frac{t_k}{\mu^{b}}\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{equation}
The derivative wrt to $\eqp^{b}_{s}$ requires a little more care (and is infact not correct as the commutator needs to be accounted for...):
\begin{equation}
	\frac{\partial Q^b_{ij}}{\partial \eqp^b_s} = -\delta_{ij}\sum_{k\neq i} \frac{\partial \eqp^b_k}{\partial \eqp^b_s} W_{ik} + (1-\delta_{ij})\delta_{is}W_{sj} = -\delta_{ij}(1-\delta_{is}) W_{is} + (1-\delta_{ij})\delta_{is}W_{sj}
\end{equation}
Using this, we can now write
\begin{equation}
\begin{split}
 \frac{\partial \log \LH}{\partial \eqp^{b}_s} &= \sum_{k\in T} t_k\mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i (-\delta_{ij}(1-\delta_{is}) W_{is} + (1-\delta_{ij})\delta_{is}W_{sj})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_s}{\eqp^b_s} \\
& =\sum_{k\in T} t_k\mu_b \frac{-\sum_{h,i\neq s}\sigma^{k,b}_i W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,j\neq s}\sigma^{k,b}_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_s}{\eqp^b_s}
\end{split}
\end{equation}
where $\sigma^{0,b}_s$ corresponds to the distribution of states of the root sequence at position $b$.
The corresponding derivative wrt to the transition matrix $W_{rs}$ has to account for the fact that $W_{ij}=W_{ji}$ and hence
\begin{equation}
\begin{split}
\frac{\partial Q^b_{ij}}{\partial W_{rs}} & = \mu^b(1-\delta_{rs})(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \mu_b\delta_{ij}(\sum_{h\neq i} p^b_h (\delta_{ir}\delta_{hs} + \delta_{is}\delta_{hr}) ) \\
& = \mu^b(1-\delta_{rs})\left[(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})\right]
\end{split}
\end{equation}
We will not consider the case $r=s$ and drop the $1-\delta_{rs}$ condition.
\begin{equation}
\begin{split}
	\frac{\partial \log \LH}{\partial W_{rs}} & = \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i \left[(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})\right] e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \\
	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i (1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h
	- \sum_{i,j,h}\sigma^{k,b}_i\delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}\\
	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{h}\left[\sigma^{k,b}_r p^b_r e^{\mat{Q}^{b}t_k}_{sh} + \sigma^{k,b}_s p^b_s e^{\mat{Q}^{b}t_k}_{rh}\right] \rho^{k,b}_h
	- \sum_{j,h}\sigma^{k,b}_j(p^b_s \delta_{jr} + p^b_r\delta_{js})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}\\
	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{h}\left[\sigma^{k,b}_r p^b_r e^{\mat{Q}^{b}t_k}_{sh} + \sigma^{k,b}_s p^b_s e^{\mat{Q}^{b}t_k}_{rh}\right] \rho^{k,b}_h
	- \sum_{h} \left[\sigma^{k,b}_r p^b_s e^{\mat{Q}^{b}t_k}_{rh} + \sigma^{k,b}_s p^b_r e^{\mat{Q}^{b}t_k}_{sh}\right] \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{split}
\end{equation}
At an extremum, all these derivatives have to vanish -- after accounting for additional constraints due to normalization.
In particular, we would like to enforce $\sum_i \eqp^b_i=0$ which results in the extramal condition
\begin{equation}
	0 = \eqp^b_s \sum_{k\in T} t_k \frac{-\sum_{h,i\neq s}\sigma^{k,b}_i W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,j\neq s}\sigma^{k,b}_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \sigma^{0,b}_s + \lambda^b \eqp^b_s
\end{equation}
To determine $\lambda^b$, we sum this over $s$.
\begin{equation}
\begin{split}
	0 & = \sum_{k\in T} t_k \frac{-\sum_{h,s,i\neq s}\sigma^{k,b}_i \eqp^b_s W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,s, j\neq s}\sigma^{k,b}_s \eqp^b_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
	0 & = \sum_{k\in T} t_k \frac{\sum_{h,s,j\neq s} \left[-\sigma^{k,b}_j \eqp^b_s W_{js} + \sigma^{k,b}_s \eqp^b_s W_{sj}\right] e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
	0 & = \sum_{k\in T} t_k \frac{\sum_{h,s,j\neq s} \left[-\sigma^{k,b}_j + \sigma^{k,b}_s\right]  \eqp^b_s W_{sj}e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j,s\neq j} \left[-\sigma^{k,b}_j\eqp^b_s W_{sj} + \sigma^{k,b}_s\eqp^b_s W_{sj}\right]  e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b\\
	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j} \left[\sigma^{k,b}_jQ_{jj} + \sum_{s\neq j}\sigma^{k,b}_sQ_{sj}\right]  e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b\\
	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j,i} \sigma^{k,b}_i Q_{ij} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b
\end{split}
\end{equation}
The first term vanishes accoring to the extremal condition for $\mu^b$ and hence $\lambda_b=-1$.
The other Lagrange parameter for $W_{ij}$ vanishes (I think) because there is no source associated with it.

There is a general pattern here: All conditions consist of a positive and negative term that need to balance.
The former is dominated by `off-diagonal' entries, i.e., when the overlap between $\sigma_i$ and $\rho_i$ is small, the latter by diagonal entries.
The ratio of the two can be cast into a multiplicative update rule for the rates.

To arrive at an iterative scheme to solve these equations, it is helpful to split the individuals terms into \emph{diagonal} contributions (the case in which the sequence does not change along a branch) and \emph{off-diagonal} contributions (substitutions happened on that branch).
We will first consider the limit of short branches were we can approximate $e^{\mat{Q}^at}=\mat{I} + \mat{Q}^at$ and evaluate to leading order
\begin{equation}
\begin{split}
t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \approx &	t_k \frac{-\sum_{i} \mu^b\sum_h p_h^b W_{hi} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i, j\neq i} \mu^b\sigma^{k,b}_i p_i^bW_{ij} \rho^{k,b}_j}{\sum_{i} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i,j\neq i} \mu^b \sigma^{k,b}_i \eqp_i^b W_{ij}t_k \rho^{k,b}_j}\\
\end{split}
\end{equation}
In the limit of short branches (or if ancestral states are known), $\sigma^{k,b}_i$ and $\rho^{k,b}_i$ are either $0$ or $1$ such that only one of the two terms in the numerator and denominator contribute.
In case $\sigma^{k,b}_i=\rho^{k,b}_i$ (no substitution on that branch), the denominator is 1 and the numerator $- \mu^b \sum_{i,h}p_h^b W_{hi}t_k$.
If a substitution happened, numerator and denominator cancel with result 1.
After summing over all branches, the condition for the mutation rate simplifies to
\begin{equation}
	\mu^b \sum_i \eqp_i^b W_{ij}T_j^b = \sum_{i\neq j} n^b_{ij}
\end{equation}
where $T_j^b$ is the sum of all $t_k$ were the sequence is in state $j$ and $n^b_ij$ are the number of times a mutation from state $j$ to $i$ at position $b$ is observed.

Going through the same manipulations for the conditions for optimal $\eqp^b_i$, we find
\begin{equation}
	\begin{split}
& t_k\frac{\sum_{h,j}\sigma^{k,b}_i Q^{b}_{ih} e^{\mat{Q}^{b}t}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \sigma_i^{0,b}+\eqp_i^b \\
&\approx t_k \frac{-\sum_{h}\sigma^{k,b}_i\mu^b p_h^b W_{ih} \rho^{k,b}_i + \sum_{j\neq i}\sigma^{k,b}_i \mu^b \eqp_i^b W_{ij} \rho^{k,b}_j}{\sum_{i} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i,j\neq i} \mu^b \sigma^{k,b}_i \eqp_i^b W_{ij}t_k \rho^{k,b}_j} + \sigma_i^{0,b}+\eqp_i^b
	\end{split}
\end{equation}
This can be solved for $p_i^{b}$ as follows
% \begin{equation}
% 	p_i^b(\sum_{h,j\neq i}\eqp_h^b W_{ih}T_i + 1) = \sum_{j\neq i}\sigma^{k,b}_i p_i^h W_{ij} \rho^{k,b}_j}{}
% \end{equation}
